---
title: "Writing a scientific proposal"
subtitle: "Proposal writing & professional development<br>FISH 521"
author: "Mark Scheuerell"
date: "10 January 2022"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## General structure | 3 important pieces

1. Title

2. Abstract

3. Body

    - text / narrative  
    - figures / tables  
    - references  
    - budget


## General structure | Title

* Must get the attention of a wide but appropriate range of potential readers.

> - Who is the target audience? What is the funding source? Related to species, region, habitat, taxa?

> - Need not be formal but avoid getting too cute

> - Pros and cons of being specific and general

> - Consider a rough title to help think about the proposal & then refine it later


## General structure | Abstract

* As with a scientific paper, it should include elements of

    - introduction  
    - methods  
    - (anticipated) results  
    - discussion  
    
> - Focus on introduction & interpretation

> - Should convey that project is **important & do-able**


## General structure | Abstract

Q: When should you write the abstract?


## General structure | Abstract

Q: When should you write the abstract?

A: Last!


## General structure | Body of proposal

> - Introduction  
> - Objectives  
> - Materials & methods  
> - Timeline  
> - Expected results & interpretation  
> - Potential difficulties  
> - Outreach, communication & engagement  
> - Data management plan  
> - Budget & justification  
> - References


## Body of proposal | {data-background=hourglass.png data-background-size=35% data-background-position="50% 70%"}


## Body of proposal | Introduction {data-background=triangle.png data-background-size=50% data-background-position="50% 70%"}

<br>

<div class="centered">
Broadest perspective

Progressively

more

specific
</div>


## Body of proposal | Objectives

* Most critical part

    - helps the writer organize thoughts  
    - tells the reader what to expect
    

## Body of proposal | Objectives

* Most critical part

    - helps the writer organize thoughts  
    - tells the reader what to expect
    

## Body of proposal | Objectives

* Consider a two-step approach

    1. The overall goal of this study is to...  
    2. Specifically, I will test the following three hypotheses


## Body of proposal | Objectives

*  Begin with general questions

> - End with specific hypotheses/predictions

> - Avoid statements like "will be different" or "will show a pattern"


## Body of proposal | Materials & methods

* Use enough detail that a reader could essentially duplicate the study (or envision it accurately)

> - Provide important details & omit irrelevant ones

> - Sub-headings can be a useful way to organize the section (eg, "Field study", "Data analysis")


## Body of proposal | Timeline {data-background=fortune.png data-background-size=60% data-background-position="50% 70%"}



## Body of proposal | Timeline

* Outline *what* will happen & *when* it will happen

> - Should mirror elements of methods section

> - Should also include time for writing


## Body of proposal | Timeline

* Gantt charts are a great way to show a timeline

* Can use Excel or many R packages (eg, `{gantrify}`)


## Gantt chart

```{r gantt}
library("ganttrify")

dat <- data.frame(wp = LETTERS[1:4],
                  activity = c("Collect data", "Analyze data",
                               "Outreach", "Write report"),
                  start_date = c(1, 4, 7, 10),
                  end_date = c(5, 7, 9, 12))
ganttrify(project = dat,
          hide_wp = TRUE,
          size_text_relative = 1.2, 
          mark_quarters = TRUE,
          project_start_date = "2022-06",
          font_family = "Roboto Condensed")
```


## Body of proposal | Expected results & interpretation  

* If the study is planned well, you should be able to describe the possible or likely outcomes

> - Consider how each outcome will be interpreted (eg, support for the hypothesis or the predicted alternative)

> - Outcomes that cannot be interpreted suggest a vague hypothesis or weak study design 

> - Be open to the possibility that your hypothesis will not be supported, or that the data will be inconclusive


## Body of proposal | Potential difficulties  

* What could go wrong?

> - Consider contingency plans

> - Can you modify the objectives or hypotheses?


## Body of proposal | Outreach, communication & engagement  



## Body of proposal | Data management plan  



## Body of proposal | Budget & justification  



## Body of proposal | References

