---
title: "Writing a scientific proposal"
subtitle: "Proposal writing & professional development<br>FISH 521"
author: "Mark Scheuerell"
date: "10 January 2022"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## General structure | 3 important pieces

1. Title

2. Abstract

3. Body

    - text / narrative  
    - figures / tables  
    - references  
    - budget


## General structure | Title

* Must get the attention of a wide but appropriate range of potential readers.

> - Who is the target audience? What is the funding source? Related to species, region, habitat, taxa?

> - Need not be formal but avoid getting too cute

> - Pros and cons of being specific and general

> - Consider a rough title to help think about the proposal & then refine it later


## General structure | Abstract

* As with a scientific paper, it should include elements of

    - introduction  
    - methods  
    - (anticipated) results  
    - discussion  
    
> - Focus on introduction & interpretation

> - Should convey that project is **important & do-able**


## General structure | Abstract

Q: When should you write the abstract?


## General structure | Abstract

Q: When should you write the abstract?

A: Last!


## General structure | Body of proposal

> - Introduction  
> - Objectives  
> - Materials & methods  
> - Timeline  
> - Expected results & interpretation  
> - Potential difficulties  
> - Outreach & communication  
> - Data management plan  
> - Budget & justification  
> - References


## Body of proposal | {data-background=hourglass.png data-background-size=35% data-background-position="50% 70%"}


## Body of proposal | Introduction {data-background=triangle.png data-background-size=50% data-background-position="50% 70%"}

<br>

<div class="centered">
Broadest perspective

Progressively

more

specific
</div>


## Body of proposal | Objectives

Most critical part

* helps the writer organize thoughts  

> - tells the reader what to expect
    

## Body of proposal | Objectives

Consider a two-step approach

1. The overall goal of this study is to...  

2. Specifically, I will test the following three hypotheses


## Body of proposal | Objectives

* Begin with general questions

> - End with specific hypotheses/predictions

> - Avoid statements like "will be different" or "will show a pattern"


## Body of proposal | Materials & methods

* Use enough detail that a reader could essentially duplicate the study (or envision it accurately)

> - Provide important details & omit irrelevant ones

> - Sub-headings can be a useful way to organize the section (eg, "Field study", "Data analysis")


## Body of proposal | Timeline {data-background=fortune.png data-background-size=60% data-background-position="50% 70%"}



## Body of proposal | Timeline

Outline *what* will happen & *when* it will happen

> - Should mirror elements of methods section

> - Should also include time for writing


## Body of proposal | Timeline

* Gantt charts are a great way to show a timeline

* Can use Excel or many R packages (eg, `{gantrify}`)


## Gantt chart

```{r gantt}
library("ganttrify")

dat <- data.frame(wp = LETTERS[1:4],
                  activity = c("Collect data", "Analyze data",
                               "Outreach", "Write report"),
                  start_date = c(1, 4, 7, 10),
                  end_date = c(5, 7, 9, 12))
ganttrify(project = dat,
          hide_wp = TRUE,
          size_text_relative = 1.2, 
          mark_quarters = TRUE,
          project_start_date = "2022-06",
          font_family = "Roboto Condensed")
```


## Body of proposal | Expected results & interpretation  

* If the study is planned well, you should be able to describe the possible or likely outcomes

> - Consider how each outcome will be interpreted (eg, support for the hypothesis or the predicted alternative)

> - Outcomes that cannot be interpreted suggest a vague hypothesis or weak study design 

> - Be open to the possibility that your hypothesis will not be supported, or that the data will be inconclusive


## Body of proposal | Potential difficulties  

What could go wrong?

> - Consider contingency plans

> - Can you modify the objectives or hypotheses?


## Body of proposal | Outreach, communication & engagement  

Two levels engagement to consider:

1. Scientific community

2. General public


## Outreach & communication | Scientific community

* Papers and conferences are the standard means

> - Conference travel & publication charges can be costly, so make sure to explicitly address these in the budget

> - Graphical abstracts, videos, and social media are becoming increasingly popular


## Animate your science

<iframe width="560" height="315" src="https://www.youtube.com/embed/O4BF8Ef_OIg?controls=0&autoplay=1" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>


## Outreach & communication | General public

* So-called "broader impacts" strengthen the relationship between the science community and society

* This is an increasingly important aspect of grant proposals


## Broader Impacts | Evaluation criteria at NSF

What is the potential for the proposed activity to benefit society or advance desired societal outcomes?

> - Match your strategy to the target community

> - For example, if you are planning for outreach at elementary schools, you will need to package your research into products children can understand.


## Broader Impacts | Evaluation criteria at NSF

To what extent do the proposed activities suggest and explore creative, original or potentially transformative concepts?

> - Describe how you will expand on known, effective strategies and give examples of successful ones

> - For example, hands-on, project-based learning enhances student performance, increases motivation and engagement, and develops critical thinking.


## Broader Impacts | Evaluation criteria at NSF

Is the plan for carrying out the proposed activities well-reasoned, well-organized and based on sound rationale?

> - Develop your Broader Impacts plan in the same way you develop the rationale for your research activities, including an evaluation plan.

> - For example, how will you gauge the success of engaging an elementary school audience?


## Broader Impacts | Evaluation criteria at NSF

How well qualified is the individual / team / institution to conduct the proposed activities?

> - Play to your strengths

> - For example, if you have a talent for creating games, perhaps you could create a game about a food web that exemplifies your research on the interaction of different fish species.

> - Think outside the box

> - For example, perhaps there is someone in the education department that would make a good collaborator.


## Broader Impacts | Evaluation criteria at NSF

Are there adequate resources available to the principal investigator to carry out the proposed activities?

> - Make sure to describe what resources you have

> - Make sure to budget for broader impacts activities


## Body of proposal | Data management plan  



## Body of proposal | Budget & justification  



## Body of proposal | References

